---
title: "Using MSPrep"
author: 
    - name: Max McGrath
      email: max.mcgrath@ucdenver.edu
      affiliation: University of Colorado Anschutz Medical Campus
    - name: Matt Mulvahill
      email: matt.mulvahill@gmail.com
      affiliation: Charter Communications
    - name: Grant Hughes
      email: dydxhughes@gmail.com
    - name: Sean Jacobson
      email: jacobsons@njhealth.org
      affiliation: National Jewish Hospital
    - name: Harrison Pielke-Lombardo
      email: harrison.pielke-lombardo@cuanschutz.edu
      affiliation: University of Colorado Anschutz Medical Campus
    - name: Katerina Kechris
      email: katerina.kechris@cuanschutz.edu
      affiliation: University of Colorado Anschutz Medical Campus
package: MSPrep
output: 
  BiocStyle::html_document:
    highlight: "tango"
    code_folding: show
    toc: true
    toc_float: 
      collapsed: false
vignette: |
  %\VignetteIndexEntry{Using MSPrep}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r, include=FALSE, echo=FALSE}
# date: "`r doc_date()`"
# "`r pkg_ver('BiocStyle')`"
# <style>
#     pre {
#     white-space: pre !important;
#     overflow-y: scroll !important;
#     height: 50vh !important;
#     }
# </style>
```

# Introduction

MSPrep provides six key functions for preparing metabolomics data for analysis. This vignette will provide:

- An explanation of each function provided by MSPrep
- An explanation of the data format required by MSPrep
- Two examples for each function demonstrating varying usage
- Notes on function usage

# Loading libraries

```{r}
library(MSPrep)
```

# Expected Format

We first need to format the data. The following format is expected for most types of data generated from mass spectrometry.

Most often, two columns in the data will specify mass-to-charge ratio and retention-time. These columns may have any names (e.g. "mz" and "rt" or "Mass" and "Retention.Time") which should be specified using the `mz` and `rt` parameters of `ms_tidy()` and `ms_summarize()`.

As an addition to (or replacement of) the mass-to-charge ratio and retention-time columns, a column may be provided which provides compound names. The name of this column should be specified using the `met_id` parameter of `ms_tidy()` and `ms_summarize()`.

The next columns are for the other sample variables. These columns should have names that contain information about their contents which will be used later in the pipeline (e.g. SampleID). There are three pieces of information which may be present at the end of each column name, separated by a consistent separator. These are the spike, batch, and replicate ID.

As an example see the provided dataset, `msquant`.

```{r}
data("msquant")
colnames(msquant)[3]
```

Above is the third column name in `msquant`. The first part "Neutral_Operator_Dif_Pos_" will not be used, so we will assign it to `col_extra_txt`. The next value, "1x", is the spike-in value. The following value, "O1", is the batch. The remaining values, "A" and "01", are the replicate and subject IDs.

Identifiers in the datasets:

- LCMS_Run_ID = operator/replicate (A-C), subject (01-03), concentration  (1x,2x,4x)
- SubjectID   = subject (01-03), concentration  (1x,2x,4x)

With our data in this format, we can start the pipeline.

# Example One

## Tidying

The first step is to extract the information from the column names discussed above, and get it into a tidy data frame. We can do this with the ms_tidy function.

```{r}
tidied_data <- ms_tidy(msquant, 
                       mz = "mz", 
                       rt = "rt", 
                       col_extra_txt = "Neutral_Operator_Dif_Pos_", 
                       separator = "_", 
                       col_names = c("spike", "batch", "replicate", "subject_id"))
```

Note, the names chosen for col_names are arbitratry and may be altered according to your preference, but they will be used later on.

## Summarizing

This step summarizes the technical replicates if present using the following procedure for each compound in each batch.

1. If there are less than a minimum proportion of the values found among the replicates (usually one or zero), leave the value empty. Otherwise proceed.
1. Calculate the coefficient of variation between the replicates using $c_v = \frac{\sigma}{\mu}$, where $\mu$ is the mean and $\sigma$ is the standard deviation.
2. For three replicates, if the coefficient of variation is above a specified level, use the median value for the compound, to correct for the large dispersion.
3. Otherwise, use the mean value of the replicates for the compound.

If no replicates are present in the data, exclude the `replicate` argument from `ms_summarize()`. In this case, the `replicate_info` and `medians` list items of the `msprep` object returned by `ms_summarize()` will be set to `NULL`.

NOTE: In the current build, ms_summarize must be called prior to filtering, imputation, or normalization regardless of the presence or absence of replicates.

```{r}
summarized_data <- ms_summarize(tidied_data, 
                                mz = "mz", 
                                rt = "rt", 
                                replicate = "replicate", 
                                batch = "batch", 
                                groupingvars = "spike", 
                                subject_id = "subject_id", 
                                cvmax = 0.50, 
                                min_proportion_present = 1/3, 
                                missing_val = 1)
```

## Filtering

This step is straightforward but very important. Simply supply a percentage of the number of samples for which a compound needs to have data present in order to be retained in the next steps.

```{r}
filtered_data <- ms_filter(summarized_data, 
                           filter_percent = 0.8)
```

## Imputation

Next, depending on the downstream analysis, you may need to impute missing data. We provide three imputation methods:

1. half-min (half the minimum value)
2. bpca (Bayesian PCA), 
3. knn (k-nearest neighbors)

Half-min is the fastest, but may introduce bias. KNN typically takes the longest. If you choose to use KNN, you can provide a value for k. By default, KNN uses samples as neighbors. By specifiying `compoundsAsNeighbors = TRUE`, compounds will be used as neighbors instead. Note that this is significantly slower than using samples as neighbors and may take several minutes or more to run depending on the size of your dataset.

```{r}
imputed_data <- ms_impute(filtered_data, 
                          imputeMethod ="knn", 
                          k_knn = 5)
```

## Normalization

In order to make comparisons between samples, the data may need to be normalized and transformed. This step performs one of eight normalization strategies. 

1. Median
2. ComBat
3. Quantile
4. quantile + ComBat
5. median + ComBat
6. CRMN
7. RUV
8. SVA

Note that all methods involving ComBat may only be used on data that contains batches. Similarly, SVA and CRMN normalization may only be used on data sets that use spike-ins (internal standards). 

Additionally, the argument `transform` enables the transformation of data prior to normalization using a log base 10 ("log10") or base 2 ("log2") transformation. Specify "none" to bypass transformation. 

For experiments which have control compounds, a list of the column numbers containing them should be provided in the controls variable. Otherwise, simply leave the controls parameter blank or NULL. 

For RUV normalization, the `k_ruv` argument specifies the number of factors on which the data is normalized.

```{r}
normalized_data <- ms_normalize(imputed_data, 
                                normalizeMethod ="CRMN", 
                                controls = NULL,  
                                n_comp = 2, 
                                n_control = 10, 
                                transform = "log10")
```

## Returning

Throughout the above functions, an elongated version of the provided data is used. This long dataset has columns corresponding to metabolite mass-to-charge ratio, retention time, abundance, and sample features like subject, replicates, and spike-ins, and it has rows corresponding to each unique combination of the sample features. 

To return the data to a wide format with metabolites as rows and samples as columns, use the `ms_return()` function.

```{r}
returned_data <- ms_return(normalized_data)
```

## Pipeline

Usually, you will want to perform the whole pipeline. This can be done in a single statement using the `ms_prepare()` function. Simply provide the function the same arguments that you would provide to the individual functions. 

```{r}

prepared_data <- ms_prepare(msquant, 
                  mz = "mz",
                  rt = "rt",
                  col_extra_txt = "Neutral_Operator_Dif_Pos_", 
                  separator = "_", 
                  col_names = c("spike", "batch", "replicate", "subject_id"),
                  replicate = "replicate",
                  batch = "batch",
                  subject_id = "subject_id",
                  groupingvars = "spike",
                  filter_percent = .8,
                  imputeMethod = "halfmin",
                  normalizeMethod = "quantile")
```

## Printing

At any point in the pipeline after `ms_summarize()`, the `print()` function provided by `MSPrep` can be called on an `msprep` object. This will print the stage of the pipeline that the object is in, along with information pertaining to each step of the pipeline.

```{r}
print(prepared_data)
```


# Example Two

Next, we will work with the dataset `COPD_131`. The raw dataset can be found [here, at Metabolomics Workbench.](https://www.metabolomicsworkbench.org/data/DRCCMetadata.php?Mode=Project&ProjectID=PR000438) Note that only a portion of the `COPD_131` dataset is included in this package in order to limit file size and improve example performance. Generally, the number of metabolites will be far greater than the number of samples and the functions included in this package will take more time to process the data.

This dataset differs from `msquant` in several ways. First, it has a column `Compound.Name` which specifies compound names, and the mass-to-charge ratio and retention-time columns are named `Mass` and `Retention.Time` respectively. Second, this dataset does not have spike-ins or batches (but it does have technical replicates). Finally, the data has already been normalized, so we will forgo that step of the MSPrep pipeline.

## Tidying

```{r}
data("COPD_131")
strsplit(colnames(COPD_131)[4], "_")
```

As with before, we begin by tidying the data. In this dataset, we have two features present in the column names: subject and replicate number, seperated by '_'. We also have the extra text 'X' appearing before each column name. So, we will run `ms_tidy()` as follows:

```{r}
tidied_data_131 <- ms_tidy(COPD_131,
                           met_id = "Compound.Name",
                           mz = "Mass", 
                           rt = "Retention.Time",
                           col_extra_txt = "X", 
                           separator = "_",
                           col_names = c("subject_id", "replicate"))
```

## Summarizing

Next, we move on to summarizing technical replicates. This process is largely the same as before, but with different column names specified. We call `ms_summarize()` as follows:

```{r}
summarized_data_131 <- ms_summarize(tidied_data_131,
                                    met_id = "Compound.Name",
                                    mz = "Mass", 
                                    rt = "Retention.Time", 
                                    subject_id = "subject_id", 
                                    replicate = "replicate",
                                    cvmax = 0.50, 
                                    min_proportion_present = 1/3, 
                                    missing_val = 0)
```

## Filtering

Again, this process is largely the same as before, choosing a filter percentage of 0.8. So, we call `ms_filter()` as follows:

```{r}
filtered_data_131 <- ms_filter(summarized_data_131, 
                               filter_percent = 0.8)
```

## Imputing

This time we will use Bayesian PCA using three principle components to impute the missing values for the data, calling `ms_impute()` as follows:

```{r}
imputed_data_131 <- ms_impute(filtered_data_131, 
                              imputeMethod ="bpca", 
                              n_pcs = 3)
```

## Returning

Now, we want our data returned to the format in which is was provided, so we call the return function provided by MSPrep.

```{r}
returned_data_131 <- ms_return(imputed_data_131)
```


## Pipeline

As with the previous example, the above steps can be performed in a pipeline using the `ms_prepare()` function. To skip the normalization step of the pipeline, set the `normalizeMethod` paramater to "none" (note that the same can be done for imputationMethod).

```{r}
prepared_data_131 <- ms_prepare(COPD_131,
                  met_id = "Compound.Name",
                  mz = "Mass", 
                  rt = "Retention.Time",
                  col_extra_txt = "X", 
                  separator = "_",
                  col_names = c("subject_id", "replicate"),
                  subject_id = "subject_id", 
                  replicate = "replicate",
                  cvmax = 0.50, 
                  min_proportion_present = 1/3, 
                  missing_val = 0,
                  filter_percent=0.8,
                  imputeMethod ="bpca", 
                  n_pcs = 3,
                  normalizeMethod = "none")
```

## Printing

As with before, the prepared COPD_131 `msprep` object may be passed to the `print()` function provided to print its status in the pipeline, along with information pertaining to each step of the pipeline.

```{r}
print(prepared_data_131)
```

