---
title: "Using MSPrep"
author: 
    - name: Max McGrath
      email: max.mcgrath@ucdenver.edu
      affiliation: University of Colorado Anschutz Medical Campus
    - name: Matt Mulvahill
      email: matt.mulvahill@gmail.com
      affiliation: Charter Communications
    - name: Grant Hughes
      email: dydxhughes@gmail.com
    - name: Sean Jacobson
      email: jacobsons@njhealth.org
      affiliation: National Jewish Hospital
    - name: Harrison Pielke-Lombardo
      email: harrison.pielke-lombardo@cuanschutz.edu
      affiliation: University of Colorado Anschutz Medical Campus
    - name: Katerina Kechris
      email: katerina.kechris@cuanschutz.edu
      affiliation: University of Colorado Anschutz Medical Campus
package: MSPrep
output: 
  BiocStyle::html_document:
    highlight: "tango"
    code_folding: show
    toc: true
    toc_float: 
      collapsed: false
vignette: |
  %\VignetteIndexEntry{Using MSPrep}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r, include=FALSE, echo=FALSE}
# date: "`r doc_date()`"
# "`r pkg_ver('BiocStyle')`"
# <style>
#     pre {
#     white-space: pre !important;
#     overflow-y: scroll !important;
#     height: 50vh !important;
#     }
# </style>
```

# Introduction

`MSPrep` provides a convenient set of functionalities used in the pre-analytic 
processing pipeline for mass spectrometry based metabolomics data. Functions are
included for the following processes commonly performed prior to analysis of 
such data:

1. Summarization of technical replicates (if available)
2. Filtering of metabolites
3. Imputation of missing values
4. Transformation, normalization, and batch correction

The sections which follow provide an explanation of each function contained in 
`MSPrep`, those functions' respective options, and examples of the pre-analysis 
pipeline using two data sets provided in the `MSPrep` package, `MSQuant` and 
`COPD_131`.

For further information, please see *MSPrep—Summarization, normalization and 
diagnostics for processing of mass spectrometry–based metabolomic data* 
(Hughes et al., 2014) and *Pre-analytic Considerations for Mass Spectrometry-Bas
ed Untargeted Metabolomics Data* (Reinhold et al., 2019).

```{r, echo=FALSE}
library(MSPrep)
```

# Expected Data Format

Data my be input as a Data Frame or `SummarizedExperiment`.

## Data Frame

When using the functions provided by `MSPrep` on a data frame, the following 
format is expected throughout the pipeline.

Most often, two or more columns of the data frame will identify unique 
compounds. This may include columns which specify the mass-to-charge ratio, the 
retention time, or the name of each compound. Using the parameter `compVars`, 
the names of these columns should be provided to each function as a vector of 
character strings.

The remainder of the columns in the data frame should specify the respective 
abundances of each compound for each sample. It is expected that one or more 
identifying variables for each sample will be specified by the column name (e.g.
Sample ID, batch number, or replicate). Each piece of information contained in 
the column names must be separated by a consistent character not present 
anywhere else in the column name. Using the parameter `sampleVars`, the sample 
variables present in the column names should be provided to each function as a 
vector of character strings specifying the order the variables appear, and the 
parameter `separator` should identify the character which separates each sample
variable. Each column name may also include consistent non-identifying text at
the beginning of each column name. This text should be provided to each function
using the `colExtraText` parameter.

As an example see the provided data set `msquant` and its use in the pipeline
below.

## SummarizedExperiment

When using the functions provided by `MSPrep` on a `SummarizedExperiment`, it is
expected that the data will include a single `assay` of abundances, `rowData`
identifying characteristics of each metabolite, and `colData` specifying
characteristics of each sample. The parameters discussed in the previous section
may be ignored.

# Example One - Technical Data Set

```{r}
data("msquant")
colnames(msquant)[3]
```

Above is the third column name in `msquant`. The first part 
"Neutral_Operator_Dif_Pos_" will not be used in this analysis, so we will assign
it to `colExtraText` parameter. The next value, "1x", is the spike-in value. 
The following value, "O1", specifies the sample's batch. The remaining values, 
"A" and "01", specify the replicate and subject IDs. We will pass these
sample variables to each function with the `sampleVars` parameter. Finally, note
that `msquant` contains two columns, `mz` and `rt` which identify each 
compounds' mass-to-charge ratio and retention time, respectively. We will pass 
these column names to each function using the `compVars` parameter.

With our data in this format, we can start the pipeline.

## Summarizing

This step summarizes the technical replicates using the following procedure for 
each compound in each batch.

1. If there are less than a minimum proportion of the values found among the 
replicates (usually one or zero), leave the value empty. Otherwise proceed.
2. Calculate the coefficient of variation between the replicates using 
$c_v = \frac{\sigma}{\mu}$, where $\mu$ is the mean and $\sigma$ is the
standard deviation.
3. For three replicates, if the coefficient of variation is above a specified 
level, use the median value for the compound, to correct for the large 
dispersion.
4. Otherwise, use the mean value of the replicates for the compound.

If the name of variable specifying replicate in `sampleVars` for a data frame 
or the column data of a `SummarizedExperiment`, specify the name of the variable
using the `replicate` parameter.

The technical replicates in `MSQuant` are summarized below using a CV maximum of
.50 and a minimum proportion present of 1 out of 3 replicates. Note that in the 
`MSQuant` dataset, missing values are represented as '1', which is specified in 
the `missingValue` argument below. `msSummarize()` will replace these missing 
values with '0' in all instances where the summarization algorithm determines 
the values to be truly missing.

```{r}
summarizedDF <- msSummarize(msquant,
                            cvMax = 0.50,
                            minPropPresent = 1/3,
                            compVars = c("mz", "rt"),
                            sampleVars = c("spike", "batch", "replicate", 
                                           "subject_id"),
                            colExtraText = "Neutral_Operator_Dif_Pos_",
                            separator = "_",
                            missingValue = 1)
```

```{r, echo = FALSE}
head(summarizedDF)
```


## Filtering

Following the summarization of technical replicates, the data can be filtered to
only contain compounds present in a specified proportion of samples. To do so, 
the `msFilter()` function is provided. By default, `msFilter()` uses the 80% 
rule and removes compounds which are are not present in 80% of the samples.

```{r}
filteredDF <- msFilter(summarizedDF,
                       filterPercent = 0.8,
                       compVars = c("mz", "rt"),
                       sampleVars = c("spike", "batch", "subject_id"),
                       separator = "_")
```

```{r, echo = FALSE}
head(filteredDF)
```

## Imputation

Next, depending on the downstream analysis, you may need to impute missing data.
Three imputation methods are provided:

1. half-min (half the minimum value)
2. bpca (Bayesian PCA), 
3. knn (k-nearest neighbors)

Half-min imputes each missing value as one half of the minimum observed value 
for that compound. Half-min imputation performs faster than other methods, 
but may introduce bias. The BPCA algorithm, provided by the `pcaMethods` 
package, estimates the missing value by a linear combination
of principle axis vectors, with the number of principle components specified by 
the user with the `nPcs` argument. KNN uses a K-Nearest Neighbors algorithm 
provided by the `VIM` package. Users may provide 
their preferred value of k using the `kKnn` argument. By default, KNN 
uses samples as neighbors, but by specifying `compoundsAsNeighbors = TRUE`, 
compounds will be used as neighbors instead. Note that this is significantly 
slower than using samples as neighbors and may take several minutes or more to 
run depending on the size of your data set.

```{r}
imputedDF <- msImpute(filteredDF,
                      imputeMethod = "knn",
                      compVars = c("mz", "rt"),
                      sampleVars = c("spike", "batch", "subject_id"),
                      separator = "_",
                      returnToSE = FALSE,
                      missingValue = 0)
```

```{r, echo = FALSE}
head(imputedDF)
```

## Normalization

In order to make comparisons between samples, the data may need to be 
transformed and normalized This step transforms the data and performs one of 
eight normalization strategies:

1. Median
2. ComBat
3. Quantile
4. Quantile + ComBat
5. Median + ComBat
6. CRMN
7. RUV
8. SVA

`msNormalize()` also provides options to transform the data using a log base 10
(default) or log base 2 transformation. To select either option, or to forego 
transformation, use the `transform` argument to specify `"log10"`, `"log2"`, or
`"none"` respectively. 

Quantile normalization, provided by the `preprocessCore` package, ensures that
the provided samples have the same quantiles. Median normalization subtracts
the median abundance of each sample from every compound in that sample, thereby
aligning the median abundance of each sample at 0.

ComBat, provided by the `sva` package, is an empirical Bayes batch effect
correction algorithm to remove unwanted batch effects and may be used separately
or in conjunction with quantile or median normalization. When using ComBat, a 
`sampleVar` called "batch" must be present for data frames, or for 
`SummarizedExperiment` "batch" must be present in the columns names of 
`colData`. Or, if the sample variable corresponding to batch differs from 
"batch", you may specify the batch variables name using the `batch` parameter.

RUV and SVA normalization each estimate a matrix of unobserved factors of 
importance using different methods of supervised factor analysis. For both 
methods, known covariates should be provided using the `covariatesOfInterest` 
parameter, and must correspond to the sample variables specified by `sampleVars`
in the case of a data frame or `colData` in the case of a 
`SummarizedExperiment`. For RUV normalization, the `kRUV` argument specifies the
number of factors on which the 
data is normalized.

Cross-Contribution Compensating Multiple Standard Normalization (CRMN), provided
by the `crmn` package, normalizes based on internal standards. The sample 
variable identifying internal standards must be provided using the
`covariatesOfInterest` parameter. For experiments which have control compounds, 
a vector of the row numbers containing them should be provided in the 
controls variable. If a vector of control compounds is not provided, data driven
controls will be generated.

Below, we apply a log base 10 transformation, quantile normalization, and ComBat
batch correction.

```{r, message = FALSE, results = 'hide'}
normalizedDF <- msNormalize(imputedDF,
                            normalizeMethod = "quantile + ComBat",
                            transform = "log10",
                            compVars = c("mz", "rt"),
                            sampleVars = c("spike", "batch", "subject_id"),
                            covariatesOfInterest = c("spike"),
                            separator = "_")
```

```{r, echo = FALSE}
head(normalizedDF)
```

## Pipeline

Often, all the above steps will need to be conducted. This can be done in a 
single statement using the `msPrepare()` function. Simply provide the 
function the same arguments that you would provide to the individual functions. 

```{r, message = FALSE, results = 'hide'}
preparedDF <- msPrepare(msquant,
                        minPropPresent = 1/3,
                        missingValue = 1,
                        filterPercent = 0.8,
                        imputeMethod = "knn",
                        normalizeMethod = "quantile + ComBat",
                        transform = "log10",
                        covariatesOfInterest = c("spike"),
                        compVars = c("mz", "rt"),
                        sampleVars = c("spike", "batch", "replicate", 
                                       "subject_id"),
                        colExtraText = "Neutral_Operator_Dif_Pos_",
                        separator = "_")
```

```{r, echo = FALSE}
head(preparedDF)
```

# Example Two - Biological Data Set

Next, the functionality of `MSPrep` will be demonstrated using the included data
`COPD_131`. The raw data set can be found [here, at Metabolomics Workbench.](https://www.metabolomicsworkbench.org/data/DRCCMetadata.php?Mode=Project&ProjectID=PR000438) 
Note that only a portion of the compounds in the original `COPD_131` data set 
are included in this package in order to limit file size and example run time.
Generally, the number of compounds in a data set will greatly 
exceed the number of samples, and the functions included in this package will 
take more time to process the data.

This data set differs from `msquant` in several ways. First, it has a column 
`Compound.Name` which specifies compound names, and the mass-to-charge ratio and
retention-time columns are named `Mass` and `Retention.Time` respectively. 
Second, this data set does not have spike-ins or batches (but it does have 
technical replicates). Finally, the data has already been normalized so that 
step of the pipeline will be excluded.

## Summarizing

Next, the technical replicates in `COPD_131` need to be summarized. This process
is largely the same as before, but with different column names passed to 
`compVars`, so `msSummarize()` is called as follows:

```{r}
summarizedSE131 <- msSummarize(COPD_131,
                               cvMax = 0.5,
                               minPropPresent = 1/3,
                               replicate = "replicate",
                               compVars = c("Mass", "Retention.Time", 
                                            "Compound.Name"),
                               sampleVars = c("subject_id", "replicate"),
                               colExtraText = "X",
                               separator = "_",
                               returnToSE = TRUE)

```

```{r, echo = FALSE}
#head(assay(summarizedSE131))
```

## Filtering

Again, this process is largely the same as before, choosing a filter percentage 
of 0.8. So, we call `msFilter()` as follows:

```{r}
filteredSE131 <- msFilter(summarizedSE131,
                          filterPercent = 0.8)
```

```{r, echo = FALSE}
#head(assay(filteredSE131))
```

## Imputing

For this example, `msImpute()` will be called using Bayesian PCA using three 
principle components to impute the missing values for the data.

```{r}
imputedSE131 <- msImpute(filteredSE131,
                         imputeMethod = "bpca",
                         nPcs = 3,
                         missingValue = 0)
```

```{r, echo = FALSE}
#head(assay(imputedSE131))
```

## Pipeline

As with the previous example, the above steps can be performed in a pipeline 
using the `msPrepare()` function. To skip the normalization step of the 
pipeline, set the `normalizeMethod` parameter to "none" (note that the same can 
be done for `imputationMethod`).

```{r}
preparedSE <- msPrepare(COPD_131,
                        cvMax = 0.5,
                        minPropPresent = 1/3,
                        compVars = c("Mass", "Retention.Time", 
                                     "Compound.Name"),
                        sampleVars = c("subject_id", "replicate"),
                        colExtraText = "X",
                        separator = "_",
                        filterPercent = 0.8,
                        imputeMethod = "bpca",
                        normalizeMethod = "none",
                        nPcs = 3,
                        missingValue = 0,
                        returnToSE = TRUE)
```

```{r, echo = FALSE}
#head(assay(imputedSE131))
```

# Session Info

```{r}
sessionInfo()
```

